---
title: "Modeling Exoplanet Counts"
author: "James Brainard"
subtitle: "subtitle"
format: 
  html:
    embed-resources: true
---

```{r setup, include=FALSE}
# Would include this chunk at beginning of report.
# Prevents code, warnings (important when loading tidyverse), and messages from outputting
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
```


```{r}
data <- read.csv("data/exoplanet_data.csv", skip = 322)
vars <- c("sy_snum", "st_mass", "st_rad", "st_lum", "st_met", "pl_orbper", "pl_orbsmax", "pl_orbeccen", "st_age", "pl_insol", "st_logg", "st_dens", "st_vsin")
data <- data %>%
  filter(pl_orbper < 10000)
```

Describing variables:

+---------------+-----------------------------------------------------------------------------+
| Variable      | Description                                                                 |
+===============+=============================================================================+
| `sy_snum`         | Number of stars in the system                                   |
+---------------+-----------------------------------------------------------------------------+
| `st_mass`         | Stellar mass                                                         |
+---------------+-----------------------------------------------------------------------------+
| `st_rad`          | Stellar radius                                                           |
+---------------+-----------------------------------------------------------------------------+
| `st_lum`          | Stellar luminosity                                         |
+---------------+-----------------------------------------------------------------------------+
| `st_met`         | Stellar metallicity                                                         |
+---------------+-----------------------------------------------------------------------------+
| `pl_orbper`          | Orbital period                                                           |
+---------------+-----------------------------------------------------------------------------+
| `pl_orbsmax`          | Semi-major axis                                         |
+---------------+-----------------------------------------------------------------------------+
| `st_mass`         | Stellar mass                                                         |
+---------------+-----------------------------------------------------------------------------+
| `pl_orbeccen`          | Eccentricity                                                           |
+---------------+-----------------------------------------------------------------------------+
| `st_age`         | Stellar age (in gigayears, or 1 billion years)                          |
+---------------+-----------------------------------------------------------------------------+
| `pl_insol`         | Insolation flux                         |
+---------------+-----------------------------------------------------------------------------+
| `st_logg`         | Stellar surface gravity                     |
+---------------+-----------------------------------------------------------------------------+
| `st_dens`         | Stellar density                        |
+---------------+-----------------------------------------------------------------------------+
| `st_vsin`         | Stellar rotational velocity                        |
+---------------+-----------------------------------------------------------------------------+

```{r}
data_long <- data %>%
  select(all_of(vars)) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value")

ggplot(data_long, aes(x = Value)) +
  geom_histogram(fill = "blue", alpha = 0.5) +
  facet_wrap(~Variable, scales = "free") +
  labs(title = "Distribution of Features",
       x = "Values",
       y = "Counts") +
  theme_minimal() # I know this is a telltale chatgpt line but I actually just really like it for this

ggplot(data, aes(x = sy_pnum)) +
  geom_bar(fill = "blue", alpha = 0.5) +
  labs(title = "Distribution of sy_pnum (Number of Planets)",
       x = "Values",
       y = "Counts") +
  theme_minimal()
```

```{r}
ggplot(data, aes(x = log10(st_rad), y = st_lum)) +
  geom_point(color = "blue", alpha = 0.5) +
  geom_smooth(color = "red") +
  labs(title = "Stellar Radius vs Luminosity",
       x = "Logarithmic Stellar Radius",
       y = "Luminosity")
```

```{r}
poisson_model <- glm(sy_pnum ~ sy_snum + st_mass + st_rad + st_lum + st_met + pl_orbper + pl_orbsmax + pl_orbeccen + st_age + pl_insol + st_logg + st_dens + st_vsin, data = data, family = poisson)

summary(poisson_model)
```
```{r}
library(MASS)

negative_binomial_model <- glm.nb(sy_pnum ~ sy_snum + st_mass + st_rad + st_lum + st_met + pl_orbper + pl_orbsmax + pl_orbeccen + st_age + pl_insol + st_logg + st_dens + st_vsin, data = data)

summary(negative_binomial_model)
```

```{r}
# Model comparison, decide which is better
```

```{r}
# Feature selection on final model
```

```{r}
# Conclusion - Answer research question, areas for improvement, and future work
```


